Lossy compression of DNA sequencing quality data

### 2015-02-02 manuscript/q-compression.tex
Changed paper to manuscript, which is a more appropriate name for the directory. Expanded BibTeX file, added and revised text in the LaTeX file, and updated citations so they conform to the BibTexfile compression.bib.

The \emph{de facto} standard for storing next-generation sequencing data is the \textsc{fastq} format. A component of \textsc{fastq} format files~\cite{cock2010sanger} is information for a quantitative measure of quality for each nucleotide based on the \textsc{phred} quality value, which is defined as $Q = -10\ log_{10}\/P$~\cite{ewing1998base}. Depending on the sequencing technology these quality values can range from 0 to 93, and the American Standard Code for Information Interchange \textsc{ascii} characters 33 to 126 (with some offset) are used to represent quality values.

Quality values are typically used throughout bioinformatics pipelines.
Certain tools, such as \textsc{fastqc}~\cite{andrews2010fastqc} and Prinseq~\cite{schmieder2011quality}  provide the user with detailed descriptions of the distribution of sequence quality scores. These analyses can then be used to filter and trim the sequencing data~\cite{martin2011cutadapt}.
The processed sequencing data can then be be used in subsequent analyses.
Certain genome assemblers utilize quality values directly to produce better assemblies~\cite{gnerre2011high}. Genome assemblers work by piecing together overlapping sequences, constructing a larger resulting sequence. Short-read alignment tools, such as Bowtie 2~\cite{langmead2012fast}, use quality values to scale the mismatch penalty in a way that mismatches between poor quality bases of the query and the reference are weighted less than mismatches occurring on high quality bases. Similar to alignment, software for detecting single nucleotide polymorphisms (\textsc{snp}s) can also rely on quality values~\cite{mckenna2010genome}. Detected \textsc{snp}s with high-quality bases are more trustworthy than those with low-quality bases, particularly in low coverage regions.

Although the quality values are useful for later steps, the trade-off between storing effectively double the amount of data and the resulting effects on subsequent analyses has not been well-documented. Previous literature on sequence data compression has largely focused on lossless compression of base calls~\cite{sato2001dna,chen2000compression,chen2002dnacompress}. If a reference exists for the biological data, sequences can be compressed using offsets into the reference~\cite{kozanitis2011compressing}. Typical compression methods fail to utilize the biases inherent in existing sequencing technologies. Depending on the sequencing technology used, the base calling is more accurate in the earlier cycles, gradually losing quality in later cycles~\cite{kozanitis2011compressing}.
To this end, methods for compressing these quality sequences utilize these general quality trends. \textsc{SlimGene}~\cite{kozanitis2011compressing} is one such tool that builds a fixed-state Markov model on adjacent quality scores and uses Huffman Coding for compression based on the transition probabilities.

@@

%The quality value provides some measure of the accuracy of a given base call. A strategy to reduce the dimensionality of these quality values is to aggregate adjacent quality values together into the same bin. Taken to the extreme, a base call can be considered either good or bad. This is the motivation of our approach, deemed good/bad encoding, where we observe the distribution of quality values across the read set and find a cutoff between good and bad bases.

In one of the more extreme cases of quality value binning, we can bin a quality value into one of two categories: good or bad.
We define good/bad encoding as finding the quality value cutoff to separate between good and bad bases.
%The goal of good/bad encoding is to find the quality value cutoff to separate between 
First, we learn the distribution of quality values across all sequences.
Using this distribution, bases are marked bad if their quality value falls below a certain threshold.
We set the cutoff as the first quartile minus 1.5 * the interquartile range (IQR), which is the difference between the first and third quartile.
1.5 * IQR is the value used by Tukey's box plot~\cite{mcgill1978variations}.
The main benefit of this approach is that it is completely data-dependent. No prior assumptions of the distribution of the quality values need to be made.

@@

K-means clustering is a vector quantization method, partitioning a set of samples into K sets that minimize within-cluster sum of squares~\cite{macqueen1967some}. Using a random subset of read quality values, a compression method can use the computed cluster centers as read quality profiles. As the problem is NP-hard, we use a heuristic iterative refinement approach quickly converging to a locally optimal minimum provided by the Python package scikit-learn 0.15~\cite{scikit-learn}.

%First, the method randomly samples from the set of quality sequences to form the training set. The amount of quality sequences sampled can be adjusted by the user.
First, the method samples an adjustable number of quality sequences randomly from the file to be used as a training set. Quality sequences are represented by vectors containing their \textsc{phred} scores corresponding to each position along the read. Subsequently, K-means clustering is performed on the training set until convergence. The obtained cluster centers will be the quality profile prototypes for the dataset.

@@

\subsection{Blockwise image compression-based algorithm}

The principle of utilizing image compression algorithms on seemingly dissimilar data such as electromyography signals is not new~\cite{costa2008compression}. Sequence quality values possess some properties such as smooth overall gradients and sudden drop-offs along a sequence whose accurate and efficient representation are crucial to image compression algorithms. Our aim is therefore to reduce the sequencing quality compression problem to a very well known and extensively researched problem, leveraging the power of mature algorithms on the image compression field. The key steps of the problem reduction are:

\begin{enumerate}
    \item Take a number of quality sequences and represent them as a matrix (rows: reads, columns: base positions).
    \item Sort its rows in a way that maximizes smoothness across the matrix, e.g. apply hierarchical clustering, akin to creating a cluster heat map.
    \item Take the resulting matrix as a grayscale image with each pixel's intensity corresponding to the \textsc{phred} score it represents.
    \item Encode the image with a suitable image compression algorithm.
    \item Either store the sorted indices to recover their original order during decompression, or re-order reads corresponding to quality values instead.
\end{enumerate}

\begin{figure*}[!tpb]%figure2
\centerline{\includegraphics[width=6.9in]{png_pipeline.png}}
\caption{Image compression based approach. Quality values are loaded into a matrix in chunks (256 reads of read-length 101 in the example). Reads are hierarchically clustered, and the rows of the matrix accordingly re-ordered. The new matrix, given its properties such as smoothness and distinct spatial features, lends itself to efficient image compression. Note the uniform dip in quality around position 35: the sequencing process was possibly temporarily compromised, affecting several thousands of reads.}\label{fig:png_method}
\end{figure*}

Lossy image compression algorithms are prone to introducing various artifacts, such as ringing, blockiness, etc. Their implication in this context is unpredictable distortion across multiple positions and even reads. Whether this is an acceptable compromise or not primarily depends on the use case.  In this study, we applied a hybrid approach using lossless Portable Network Graphics (\textsc{png}) encoding on adaptively quantized data, all performed by the \textsc{png} encoder pngquant 2.3.0 (http://pngquant.org/). This method is free of the previously mentioned artifacts, and provides an adjustable level of compression.

The choice of this particular image encoder is not essential to the algorithm, but rather serves as an initial proof of concept implementation. Other image compression algorithms could be explored extensively, and seamlessly substituted into our method for further benchmarking.

Our method uses mean linkage hierarchical clustering for row sorting, which, just like the image encoder, can be replaced with other linkage criteria or even sorting methods to be evaluated for performance.

@@

For our analyses, we use sequencing data provided by the \textsc{GAGE} project~\cite{salzberg2012gage}.
The read data for \textit{Rhodobacter sphaeroides} 2.4.1 was downloaded from
\href{http://gage.cbcb.umd.edu/data/Rhodobacter_sphaeroides}{http://gage.cbcb.umd.edu/data/Rhodobacter\_sphaeroides},
%http://gage.cbcb.umd.edu/data/Rhodobacter\_sphaeroides
and the corresponding reference sequence was obtained from the NCBI RefSeq database (NC\_007488.1, NC\_007489.1, NC\_007490.1, NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1).
The \textit{Rhodobacter sphaeroides} 2.4.1 data set consists of 2,050,868 paired-end reads (insert size of 180 bp) and 2,050,868 short-jump reads (insert size of 3,000 bp).

%\begin{table}[!t]
%\processtable{This is table caption\label{Tab:01}}
%{\begin{tabular}{llll}\toprule
%head1 & head2 & head3 & head4\\\midrule
%row1 & row1 & row1 & row1\\
%row2 & row2 & row2 & row2\\
%row3 & row3 & row3 & row3\\
%row4 & row4 & row4 & row4\\\botrule
%\end{tabular}}{This is a footnote}
%\end{table}

@@

For quality control, we use Autoadapt tool which uses \textsc{fastqc}~\cite{andrews2010fastqc} and \textsc{CutAdapt}~\cite{martin2011cutadapt} packages with the default settings. PrinSeq~\cite{schmieder2011quality} was used to count the number of bases and reads in each \textsc{fastq} file.

For \emph{de novo} genome assembly, we use \textsc{allpaths-lg}~\cite{gnerre2011high} version r50191 with default settings and 32 threads.

For aligning the sequences, we use Bowtie 2~\cite{langmead2012fast} with default settings and modified max- and min-mismatch parameters detailed below.

@@

Genome assemblers, such as \textsc{allpaths-lg}~\cite{gnerre2011high}, often utilize quality values to aid in the assembly process. By reducing the information contained within the quality values, we introduce the potential for the resulting assembly to be affected.

\textsc{allpaths-lg} is one of the leading bacterial genome assemblers~\cite{salzberg2012gage}. We examine the effect of our lossy compression methods on the \textit{Rhodobacter sphaeroides} 2.4.1 data set provided by \textsc{gage} (Genome Assembly Gold-Standard Evaluations)~\cite{salzberg2012gage}. \textsc{allpaths-lg} is used to assemble the original data along with the quality compressed versions using good/bad encoding, 0--5th-order polynomial regression encodings, 32- and 256-profile encodings, and \textsc{png} encoding. We evaluate the assembly based on contiguity statistics, log average read probability (LAP~\cite{ghodsi2013novo}), and a collection of reference-based metrics. The contiguity statistics include number of scaffolds and NG50, which is defined as the weighted median contig size (the length of largest contig $c$ such that the total size of the contigs larger than $c$ exceeds half of the known genome size).  The LAP score can be viewed as a log likelihood score, where a value closer to 0 is better. We use a reference-based evaluation script provided by \textsc{gage} which counts single nucleotide polymorphisms (\textsc{snp}s), relocations, translations, and inversions. The reference-based metrics are normalized by the length of the assembly to aid in comparison. Alongside the raw metrics, we also provide a cummulative sum of the individual ranks of each method for each metric. A similar approach was utilized by the \textsc{Assemblathon1} competition~\cite{earl2011assemblathon}.

@@

It is important to highlight the result that the 4th-order polynomial regression outperforms the 5th-order in all of the reference-based metrics, but not the \emph{de novo} LAP metrics. The 5th-order polynomial regression assembles approximately 2.5 kb more sequence and uses roughly $1.2\%$ more of the total read set than the 4th-order (Supplemental Table 1), which may explain the discrepency in LAP values because the LAP value is influenced by total coverage~\cite{ghodsi2013novo}.

\begin{table*}[!tpb]
%\processtable{Rankings of different assemblies created using proposed compression methods. Assemblies were created using \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides} 2.4.1. data set provided by \textsc{gage}\citep{salzberg2012gage}.\label{tab:assembly_ranks}}
\centering
\caption[]{Rankings of different assemblies produced using proposed compression methods. Assemblies were constructed using \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides} 2.4.1. data set provided by \textsc{gage}~\cite{salzberg2012gage}. Regression methods are sorted by their overall rank (determined by the sum of individual rankings).}
{\begin{tabular}{lcccccccc}
& Overall & & & & & & LAP & LAP \\
Compression            & rank & \textsc{snp}s & Inversions & Relocations & Translocations & Indels & fragment & short-jump \\
\hline
None                   & 1    & 5    & 1          & 1           & 1              & 1      & 1          & 1                    \\
PNG                    & 2    & 4    & 1          & 2           & 2              & 3      & 4          & 2                    \\
Profile (256)          & 3    & 2    & 1          & 3           & 3              & 4      & 2          & 4                    \\
Profile (32)           & 4    & 3    & 1          & 4           & 6              & 5      & 3          & 3                    \\
Regression (4th order) & 5    & 6    & 1          & 6           & 5              & 2      & 7          & 8                    \\
Regression (5th order) & 6    & 7    & 1          & 5           & 7              & 7      & 5          & 7                    \\
Good/bad               & 7    & 1    & 1          & 8           & 11             & 10     & 6          & 5                    \\
Regression (3rd order) & 8    & 8    & 1          & 9           & 9              & 6      & 8          & 6                    \\
Regression (2nd order) & 9    & 9    & 1          & 7           & 4              & 8      & 9          & 10                   \\
Regression (1st order) & 10   & 10   & 11         & 10          & 8              & 9      & 10         & 9                    \\
Regression (0 order) & 11   & 11   & 10         & 11          & 10             & 11     & 11         & 11
\\ \hline
\end{tabular}
}
\label{tab:assembly_ranks}
\end{table*}

@@

From an economics perspective, we can examine the fixed and marginal costs associated with our lossy compression techniques with respect to memory and computation.
For our quality value binning, modeling-based, and image-based compression methods there are no explicit fixed costs in terms of memory, unlike the profile--based approach, which has a small fixed cost in storing the profiles.
However, the profile-based approach has a much smaller marginal compared to the rest due to the fact that we are simply storing the profile index for each quality sequence.

@@

We have simply provided here the initial steps in analyzing the effect of lossy compression on quality sequences using a single, high-coverage bacterial data set. More work needs to be done using additional biological data sets, such as human and mouse, along with different sequencing technologies.
A more direct comparison against related lossy compression tools, such as \textsc{SlimGene}~\cite{kozanitis2011compressing} and \textsc{QualComp}~\cite{ochoa2013qualcomp}, needs to be performed. Additionally, other types of sequencing data can be analyzed apart from the Illumina data examined here. For example, the PacBio sequencing instruments produce very long reads (with average read lengths on the order of 10 kbp), but with the trade-off of having a high error-rate ($\sim$15\%).
Unlike the class of quality values we have examined here, the distribution of erroneous bases is relatively uniform~\cite{ferrarini2013evaluation}.
Koren et al.~\cite{koren2013reducing} have shown that the assembly complexity of bacterial genomes can be greatly simplified, producing near complete genome assemblies, by utilizing a single run of these long reads.
If long read sequencing technologies such as PacBio become more widely adopted, it would be of huge benefit to examine the potential of lossy compression algorithms on not only the quality values, but the biological sequencing data themselves.

@@

%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
\bibliographystyle{plain}
%
\bibliography{Compression}

>>>>

The \emph{de facto} standard for storing next-generation sequencing
data is the \textsc{fastq} format. A component of \textsc{fastq}
format files~\cite{Cock:2010ve} is information for a quantitative
measure of quality for each nucleotide based on the \textsc{phred}
quality value, which is defined as $Q =
-10\ log_{10}\/P$~\cite{Ewing:1998zr,Ewing:1998ly}. Depending on the
sequencing technology these quality values can range from 0 to 93, and
the American Standard Code for Information Interchange \textsc{ascii}
characters 33 to 126 (with some offset) are used to represent quality
values.

Quality values are typically used throughout bioinformatics pipelines.
Certain tools, such as \textsc{fastqc}~\cite{citeulike:11583827} and
Prinseq~\cite{Schmieder:2011gd} provide the user with detailed
descriptions of the distribution of sequence quality scores. These
analyses can then be used to filter and trim the sequencing
data~\cite{EJ200}. The processed sequencing data can then be be used
in subsequent analyses. Certain genome assemblers utilize quality
values directly to produce better
assemblies~\cite{Gnerre:2011kx}. Genome assemblers work by piecing
together overlapping sequences, constructing a larger resulting
sequence. Short-read alignment tools, such as Bowtie
2~\cite{Langmead:2012rw}, use quality values to scale the mismatch
penalty in a way that mismatches between poor quality bases of the
query and the reference are weighted less than mismatches occurring on
high quality bases. Similar to alignment, software for detecting
single nucleotide polymorphisms (\textsc{snp}s) can also rely on
quality values~\cite{McKenna:2010bh}. Detected \textsc{snp}s with
high-quality bases are more trustworthy than those with low-quality
bases, particularly in low coverage regions.

Although the quality values are useful for later steps, the trade-off
between storing effectively double the amount of data and the
resulting effects on subsequent analyses has not been
well-documented. Previous literature on sequence data compression has
largely focused on lossless compression of base
calls~\cite{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Giancarlo:2014rw}.
If a reference exists for the biological data, sequences can be
compressed using offsets into the
reference~\cite{Kozanitis:2011kl}. Typical compression methods fail to
utilize the biases inherent in existing sequencing
technologies. Depending on the sequencing technology used, the base
calling is more accurate in the earlier cycles, gradually losing
quality in later cycles~\cite{Kozanitis:2011kl}. To this end, methods
for compressing these quality sequences utilize these general quality
trends. \textsc{SlimGene}~\cite{Kozanitis:2011kl} is one such tool
that builds a fixed-state Markov model on adjacent quality scores and
uses Huffman Coding for compression based on the transition
probabilities.

@@

%The quality value provides some measure of the accuracy of a given
%base call. A strategy to reduce the dimensionality of these quality
%values is to aggregate adjacent quality values together into the same
%bin. Taken to the extreme, a base call can be considered either good
%or bad. This is the motivation of our approach, deemed good/bad
%encoding, where we observe the distribution of quality values across
%the read set and find a cutoff between good and bad bases.

In one of the more extreme cases of quality value binning, we can bin
a quality value into one of two categories: good or bad. We define
good/bad encoding as finding the quality value cutoff to separate
between good and bad bases. First, we learn the distribution of
quality values across all sequences. Using this distribution, bases
are marked bad if their quality value falls below a certain
threshold. We set the cutoff as the first quartile minus 1.5 * the
interquartile range (IQR), which is the difference between the first
and third quartile. 1.5 * IQR is the value used by Tukey's box
plot~\cite{mcgill1978variations}. The main benefit of this approach is
that it is completely data-dependent. No prior assumptions of the
distribution of the quality values need to be made.

@@

K-means clustering is a vector quantization method, partitioning a set
of samples into K sets that minimize within-cluster sum of
squares~\cite{macqueen1967some}. Using a random subset of read quality
values, a compression method can use the computed cluster centers as
read quality profiles. As the problem is NP-hard, we use a heuristic
iterative refinement approach quickly converging to a locally optimal
minimum provided by the Python package scikit-learn
0.15~\cite{pedregosa2011scikit}.

First, the method samples an adjustable number of quality sequences
randomly from the file to be used as a training set. Quality sequences
are represented by vectors containing their \textsc{phred} scores
corresponding to each position along the read. Subsequently, K-means
clustering is performed on the training set until convergence. The
obtained cluster centers will be the quality profile prototypes for
the dataset.

@@

@@

For our analyses, we use sequencing data provided by the \textsc{GAGE}
project~\cite{Salzberg:2012rc}. The read data for \textit{Rhodobacter
  sphaeroides} 2.4.1 was downloaded from
\href{http://gage.cbcb.umd.edu/data/Rhodobacter_sphaeroides}{http://gage.cbcb.umd.edu/data/Rhodobacter\_sphaeroides},
and the corresponding reference sequence was obtained from the NCBI
RefSeq database (NC\_007488.1, NC\_007489.1, NC\_007490.1,
NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1). The
\textit{Rhodobacter sphaeroides} 2.4.1 data set consists of 2,050,868
paired-end reads (insert size of 180 bp) and 2,050,868 short-jump
reads (insert size of 3,000 bp).

@@

For quality control, we use Autoadapt tool which uses
\textsc{fastqc}~\cite{citeulike:11583827} and
\textsc{CutAdapt}~\cite{EJ200} packages with the default
settings. PrinSeq~\cite{Schmieder:2011gd} was used to count the number
of bases and reads in each \textsc{fastq} file.

For \emph{de novo} genome assembly, we use
\textsc{allpaths-lg}~\cite{Gnerre:2011kx} version r50191 with default
settings and 32 threads.

For aligning the sequences, we use Bowtie 2~\cite{Langmead:2012rw}
with default settings and modified max- and min-mismatch parameters
detailed below.

@@

Genome assemblers, such as \textsc{allpaths-lg}~\cite{Gnerre:2011kx},
often utilize quality values to aid in the assembly process. By
reducing the information contained within the quality values, we
introduce the potential for the resulting assembly to be affected.

\textsc{allpaths-lg} is one of the leading bacterial genome
assemblers~\cite{Salzberg:2012rc}. We examine the effect of our lossy
compression methods on the \textit{Rhodobacter sphaeroides} 2.4.1 data
set provided by \textsc{gage} (Genome Assembly Gold-Standard
Evaluations)~\cite{Salzberg:2012rc}. \textsc{allpaths-lg} is used to
assemble the original data along with the quality compressed versions
using good/bad encoding, 0--5th-order polynomial regression encodings,
32- and 256-profile encodings, and \textsc{png} encoding. We evaluate
the assembly based on contiguity statistics, log average read
probability (LAP~\cite{Ghodsi:2013hb}), and a collection of
reference-based metrics. The contiguity statistics include number of
scaffolds and NG50, which is defined as the weighted median contig
size (the length of largest contig $c$ such that the total size of the
contigs larger than $c$â exceeds half of the known genome size). The
LAP score can be viewed as a log likelihood score, where a value
closer to 0 is better. We use a reference-based evaluation script
provided by \textsc{gage} which counts single nucleotide polymorphisms
(\textsc{snp}s), relocations, translations, and inversions. The
reference-based metrics are normalized by the length of the assembly
to aid in comparison. Alongside the raw metrics, we also provide a
cummulative sum of the individual ranks of each method for each
metric. A similar approach was utilized by the \textsc{Assemblathon1}
competition~\cite{Earl:2011fv}.

@@

It is important to highlight the result that the 4th-order polynomial
regression outperforms the 5th-order in all of the reference-based
metrics, but not the \emph{de novo} LAP metrics. The 5th-order
polynomial regression assembles approximately 2.5 kb more sequence and
uses roughly $1.2\%$ more of the total read set than the 4th-order
(Supplemental Table 1), which may explain the discrepency in LAP
values because the LAP value is influenced by total
coverage~\cite{Ghodsi:2013hb}.

\begin{table*}[!tpb]
\centering
\caption[]{Rankings of different assemblies produced using proposed
  compression methods. Assemblies were constructed using
  \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides}
  2.4.1. data set provided by
  \textsc{gage}~\cite{Salzberg:2012rc}. Regression methods are sorted
  by their overall rank (determined by the sum of individual
  rankings).}
{\begin{tabular}{lcccccccc}
& Overall & & & & & & LAP & LAP \\
Compression            & rank & \textsc{snp}s & Inversions & Relocations & Translocations & Indels & fragment & short-jump \\
\hline
None                   & 1    & 5    & 1          & 1           & 1              & 1      & 1          & 1                    \\
PNG                    & 2    & 4    & 1          & 2           & 2              & 3      & 4          & 2                    \\
Profile (256)          & 3    & 2    & 1          & 3           & 3              & 4      & 2          & 4                    \\
Profile (32)           & 4    & 3    & 1          & 4           & 6              & 5      & 3          & 3                    \\
Regression (4th order) & 5    & 6    & 1          & 6           & 5              & 2      & 7          & 8                    \\
Regression (5th order) & 6    & 7    & 1          & 5           & 7              & 7      & 5          & 7                    \\
Good/bad               & 7    & 1    & 1          & 8           & 11             & 10     & 6          & 5                    \\
Regression (3rd order) & 8    & 8    & 1          & 9           & 9              & 6      & 8          & 6                    \\
Regression (2nd order) & 9    & 9    & 1          & 7           & 4              & 8      & 9          & 10                   \\
Regression (1st order) & 10   & 10   & 11         & 10          & 8              & 9      & 10         & 9                    \\
Regression (0 order) & 11   & 11   & 10         & 11          & 10             & 11     & 11         & 11
\\ \hline
\end{tabular}
}
\label{tab:assembly_ranks}
\end{table*}

@@

From an economics perspective, we can examine the fixed and marginal
costs associated with our lossy compression techniques with respect to
memory and computation. For our quality value binning and model-based
compression methods there are no explicit fixed costs
in terms of memory, unlike the profile--based approach, which has a
small fixed cost in storing the profiles. However, the profile-based
approach has a much smaller marginal compared to the rest due to the
fact that we are simply storing the profile index for each quality
sequence.

@@

We have simply provided here the initial steps in analyzing the effect
of lossy compression on quality sequences using a single,
high-coverage bacterial data set. More work needs to be done using
additional biological data sets, such as human and mouse, along with
different sequencing technologies. A more direct comparison against
related lossy compression tools, such as
\textsc{SlimGene}~\cite{Kozanitis:2011kl} and
\textsc{QualComp}~\cite{Ochoa:2013rt}, needs to be
performed. Additionally, other types of sequencing data can be
analyzed apart from the Illumina data examined here. For example, the
PacBio sequencing instruments produce very long reads (with average
read lengths on the order of 10 kbp), but with the trade-off of having
a high error-rate ($\sim$15\%). Unlike the class of quality values we
have examined here, the distribution of erroneous bases is relatively
uniform~\cite{Ferrarini:2013vf}. Koren et al.~\cite{Koren:2013ye} have
shown that the assembly complexity of bacterial genomes can be greatly
simplified, producing near complete genome assemblies, by utilizing a
single run of these long reads. If long read sequencing technologies
such as PacBio become more widely adopted, it would be of huge benefit
to examine the potential of lossy compression algorithms on not only
the quality values, but the biological sequencing data themselves.

@@

\section{Some Alternative Text}

The following text is taken from a draft grant proposal, and the
content overlaps considerably with the other text in this
document. However, at least in some cases the wording of this
alternative text might be preferable and serve as a replacement. In
any event the entire draft manuscript has to be substantially revised.

\subsection{Background}

Sequencing read data constitutes the largest category of sequence data
in genomics research because of its great redundancy, inclusion of
quality values, and read-level naming and metadata. In recognition of
this abundance and importance we place particular emphasis on
compression of read data, which is the focus of this Aim. Among the
several challenges for compression of read data is dealing with
different error profiles resulting from the differences in underlying
chemistries, signal detection and processing mechanisms, inherent
biases, and other idiosyncratic properties of distinct high-throughput
sequencing technologies. Our proposed research includes development of
compression methods that robustly accommodate different error profiles
associated with sequence read data, and where advantageous we will
leverage error profiles to increase compression performance.

At the level of sequence read the probability of error for each
base-call is typically represented by \textsc{phred} quality value,
which is defined as $Q =
-10\,log_{10}P$~\cite{Ewing:1998zr,Ewing:1998ly}.  Depending on the
sequencing technology these quality values can range from 0 to 93, and
are represented with the \textsc{ascii} characters 33 to 126 (with
some offset). Quality values comprise a standard component of
\textsc{fastq} format files~\cite{Cock:2010ve}. The characteristics of
different sequencing technology-associated error profiles are
reflected in the quality values, and can be broadly described for both
Illumina and PacBio sequencing technologies.

Illumina reads are characterized by having far fewer insertion and
deletion errors, where substitution (miscall) errors are much more
frequent and have context-specific patterns. The errors and quality
values are non-uniformly distributed over the read length (e.g., error
rates increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and
32.8 -- 67.9\% of reads have low quality tails at the 3$^{\prime}$
end~\cite{Minoche:2011km}). Illumina reads have a single quality value
per base-call.

\subsection{Lossy compression of quality values} \label{sec:qual}

\subsubsection{Pilot Study: profile-based compression of quality 
values}\label{sec:profile-based}

As described previously, quality values in Illumina reads are
non-uniformly distributed over the read length and show a general
pattern of decreasing toward the 3$^{\prime}$ end. As large sets of
quality strings show similar trends of quality over their sequence,
one approach would be to identify common patterns in the data,
generate profiles representing these patterns, and use the profiles as
proxies for read quality value sequences. We demonstrated the efficacy
of this approach in a pilot study using test data of Illumina reads
the \textit{Rhodobacter sphaeroides} 2.4.1 data set provided by
\textsc{gage} (Genome Assembly Gold-Standard
Evaluations)~\cite{Salzberg:2012rc}, which consists of 2,050,868
paired-end reads (fragment library; insert size of 180 bp) and
2,050,868 reads (short-jump library; insert size of 3,000 bp).

Quality value patterns can be readily determined by clustering
sequences of quality values and using the resulting cluster centers as
representative profiles. In our pilot study we used $K$-means
clustering, a vector quantization method that partitions a set of
samples into $K$ sets that minimize within-cluster sum of
squares~\cite{hastie_09_elements-of.statistical-learning}, as follows.

\begin{enumerate}
\item
Randomly sample an arbitrary number of quality sequences as a training
set.
\item
Perform $K$-means clustering on the sample, and use the cluster
centers as profiles for the dataset.
\item
Replace each quality value sequence with the index value of the
profile closest in Euclidean distance.
\end{enumerate}

Our pilot study results demonstrate that $K$ values as small as
32--256 sufficiently represent patterns in Illumina read quality
values such that performance in subsequent analyses is very close to
uncompressed data (see Section \ref{sec:pilot-evaluation}). The
resulting compressed quality component of the original set of
\textsc{fastq} files consists of the $K$ quality profiles (the fixed
size cost), and a single byte representing the assigned quality
profile index for each read (the marginal size cost). Therefore the
compression factor is proportional of read length: the marginal size
cost of 1 byte/read for Illumina HiSeq read lengths of 100 bp requires
0.01 bits/base-call, and MiSeq read lengths of 300 bp require 0.0033
bits/base-call. The fixed cost of profiles of course reduces the
compression ratio, but this fixed cost is amortized across reads in a
set. Both fixed and marginal costs can be further reduced with
subsequent compression of the index and profiles (e.g., using general
compression algorithms). Our pilot study demonstrates that using 256
profiles with subsequent bzip2 compression (using a Burrows-Wheeler
block-sorting compression algorithm, and Huffman coding) represents
quality values of Illumina 100 bp reads at 0.074 bits/base-call
compared to the original data compressed with bzip2 only of 2.11
bits/base-call for the fragment library and 2.70 bits/base-call for
the short jump library. The very recently published
Read-Quality-Sparsifier
(\textsc{rqs};~\cite{DBLP:conf/recomb/YuYB14}), which apparently
represents the current state-of-the-art for compressing quality
values, achieved best case compression of 0.254 bits/base-call, and a
mean compression of 1.841 bits/base-call. Whereas our profile method
achieves $\sim$3.432$\times$ more compression than textsc{rqs} best
performance, and $\sim$24.878$\times$ more than average
performance. Note too that the compression ratio for \textsc{rqs}
varies with the $k$-mer composition of the data, whereas profile-based
compression is invariant to $k$-mer composition.

\subsubsection{Evaluating compression effects on subsequent 
analysis results}\label{sec:pilot-evaluation}

Lossy compression compromises representation accuracy for better space
efficiency. Different lossy compression methods have different
properties, therefore we need to evaluate performance in a number of
analysis contexts, and use multiple performance measures to
appropriately evaluate lossy compression methods. Important
bioinformatics contexts include typical tasks, such as quality control
processing, \textit{de novo} genome assembly, and aligning reads to a
reference. In our pilot study we again used Illumina reads from
fragment and short-jump libraries of the \textit{Rhodobacter
  sphaeroides} as test data, and assessed the effects of lossy
compression on each of the aforementioned bioinformatics
operations. The results for the profile-based compression follow.

\paragraph{Read quality processing}
Among the most fundamental uses of sequence quality values is as part
of the quality assessment and quality control (\textsc{qa/qc})
processes prior to subsequent analysis steps. Quality control based on
quality values generally includes two operations:
\textit{i}.~filtering, the elimination of reads that on the whole do
not meet quality standards, which reduces the total number of reads;
and \textit{ii}.~trimming of low quality base-calls from reads, which
reduces the number total number of bases.

To evaluate the effect of compression on these quality control
operations, we used \textsc{autoadapt}, which uses
\textsc{FastQC}~\cite{citeulike:11583827} and
\textsc{cutadapt}~\cite{EJ200} packages, all with the default
settings. \textsc{PrinSeq}~\cite{Schmieder:2011gd} was then used to
count the number of bases and reads in each quality control-processed
\textsc{fastq} file. The proportion of reads retained using
compressed/decompressed data compared to that of the uncompressed
data: 0.998 for the fragment library reads, and 1.013 for the
short-jump library reads; and proportion of bases retained 1.0 for the
fragment library reads, and 1.027 for the short-jump library
reads. Thus there is very little difference between profile-based
compressed data compared to that of the original uncompressed data in
read filtering and trimming.

\paragraph{\textit{De novo} genome assembly}
For de novo genome assembly, we use
\textsc{allpaths-lg}~\cite{Gnerre:2011kx} version r50191 with default
settings. We evaluate the assembly based on contiguity statistics, log
average read probability (\textsc{lap})~\cite{Ghodsi:2013hb}, and a
collection of reference-based metrics, of which the contiguity
statistics include number of scaffolds and NG50; the latter statistic
is defined as the weighted median contig size (the length of largest
contig $c$ such that the total size of the contigs larger than $c$
exceeds half of the known genome size). The \textsc{lap} score can be
viewed as a log likelihood score, where a larger value is better. We
use a reference-based evaluation script provided by \textsc{gage},
which counts single nucleotide polymorphisms (\textsc{snp}s),
relocations, translations, and inversions. The reference- based
metrics are normalized by the length of the assembly to aid in
comparison. The differences in assembly statistics between using the
profile-based compressed/decompressed data compared to using
uncompressed data are extremely small: total assembly size difference
of 6.745 kbp ($\sim$0.00015); NG50 size difference of 2.259 kbp
(0.0007); an \textsc{lap} score difference of 0.002 for the fragment
library and 0.04 for the short-jump library.

\paragraph{Aligning reads to a reference sequence}
Another important analysis involving read data are the alignments of
reads to a reference sequence. Such alignments are performed for
genome annotation, quantification of gene expression in RNA-Seq
studies, estimation of sequencing coverage, and other purposes. To
evaluate the effect of compression on read alignment we used
\textsc{bowtie} 2~\cite{Langmead:2012rw} with default settings and
max- and min-mismatch parameters of \texttt{-mp 6,6} and \texttt{-mp
  2,2} respectively. The proportion of reads mapped using
compressed/decompressed data was extremely close to that of the
uncompressed data: 0.999 for the fragment library reads, and 0.998 for
the short-jump library reads.

%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
\bibliographystyle{plain}
%
\bibliography{compression}
